Statistical Searches for Microlensing Events in Large, Non-Uniformly Sampled Time-Domain Surveys: A Test Using Palomar Transient Factory Data

### 2013-10-18 text/paper/ms.tex
added brink citation

\slugcomment{DRAFT \today}
\shorttitle{Microlensing \& PTF}
\shortauthors{Price-Whelan et al.}

@@

\title{Statistical Searches for Microlensing Events in Large, Non-Uniformly Sampled Time-Domain Surveys: A Test Using Palomar Transient Factory Data}
\author{Adrian~M.~Price-Whelan\altaffilmark{1,8} Marcel~A.~Ag\"ueros\altaffilmark{1}, Amanda P. Fournier\altaffilmark{2,3}, Rachel Street\altaffilmark{3}, Eran O. Ofek\altaffilmark{4}, Kevin~R.~Covey\altaffilmark{5}, David Levitan\altaffilmark{6}, Russ R.~Laher\altaffilmark{7}, Branimir Sesar\altaffilmark{6}, Jason Surace\altaffilmark{7}}

\altaffiltext{1}{Department of Astronomy, Columbia University, 550 W 120th St., New York, NY 10027, USA}
\altaffiltext{2}{Department of Physics, Broida Hall, University of California, Santa Barbara, CA 93106, USA}
\altaffiltext{3}{Las Cumbres Observatory Global Telescope Network, Inc., 6740 Cortona Dr.\ Suite 102, Santa Barbara, CA 93117, USA}
\altaffiltext{4}{Benoziyo Center for Astrophysics, Weizmann Institute of Science, 76100 Rehovot, Israel}
\altaffiltext{5}{Lowell Observatory, 1400 West Mars Hill Rd, Flagstaff, AZ 86001, USA}
\altaffiltext{6}{Division of Physics, Mathematics, and Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
\altaffiltext{7}{Spitzer Science Center, California Institute of Technology, Mail Stop 314-6, Pasadena, CA 91125, USA}

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged $\sim$40 times in \textit{R}-band, $\sim$$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal time sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio, performs best for identifying simulated microlensing events in our data. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations --- a total of $1.1\times10^9$ light curves --- to uncover three candidate sparse-field microlensing events. While these candidates cannot be confirmed, as we lack simultaneous, multi-color photometry of the events, this number is consistent with predictions for the microlensing event rate in the PTF footprint over the survey's three years of operations, as estimated from the model proposed in \citet{han2008}. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large data sets, both of which will be useful to future wide-field, time-domain surveys, such as that planned with the Large Synoptic Survey Telescope. 
\end{abstract}

@@

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005}, and to identify extrasolar planets \citep[][and references therein]{gaudi2011}. Such studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year.\footnote{http://ogle.astrouw.edu.pl/ogle4/ews/ews.html} In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects, such as low-mass stars, sub-stellar objects, and isolated neutron stars.

Microlensing surveys typically focus on high-density stellar regions, such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.,][]{original_ogle, original_macho, eros_original, crotts1996}, but microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the microlensing event rate in low-density stellar regions will be smaller, compared to events found by modern microlensing surveys, these events are expected to involve closer sources and lenses, and less crowded backgrounds \citep{mesolensing}. Thus, such lenses offer more opportunities to constrain the properties of the events.

@@

In this paper, we use the Palomar Transient Factory (PTF) survey, described in Section~\ref{sec:ptf}, as a test case for studying the frequency and detectability of microlensing events in all-sky, irregularly sampled, time-domain surveys. Although to date no microlensing event has been reported using PTF data, a simple extrapolation from previous all-sky rate estimates suggests $\sim$1--10 detectable microlensing events within the PTF footprint over the survey's three years of operation \citep{han2008, gaudi2008}. We examine the selection methods used by microlensing surveys to identify event signatures in light curves as well as an event identification procedure based on a set of variability statistics. We then present the detection efficiency for these methods as computed for a representative set of PTF fields (Section~\ref{sec:event_recovery}). In Section~\ref{sec:search}, we apply a set of criteria that use the results of these tests to all PTF light curves with $>$10 high-quality observations to identify candidate microlensing events. We discuss eight interesting transients and three candidate events identified in this manner before concluding in Section~\ref{sec:conc}. % A detailed investigation of the all-sky microlensing event rate,  generated from a model that includes realistic mass and velocity distributions, will be presented in Fournier et al. (in prep).

@@

As of 2012 Dec, the PTF footprint includes $\sim$15,224 (2766) deg$^2$ imaged $>$10 ($>$100) times in $R$-band and $\sim$5430 (290) deg$^2$ imaged that often in $g$-band (see Figure~\ref{fig:survey_footprint}). The PTF survey footprint is not uniformly sampled either spatially or temporally. Each field has a unique sampling pattern determined by which of the PTF sub-surveys it belongs to, what time of the year it is visible, and how high a priority it is given by the scheduler. The imaging cadences range from 1-5 days, but there is a higher-cadence campaign of a field in Orion to find transiting planets around young stars \citep{nick2009}. The result is that PTF light curves often contain gaps and regions of high-cadence observations and/or of low-cadence observations, so that the archive is a massive dataset of irregularly-sampled, time-domain photometry. Figure~\ref{fig:sampling} shows six randomly selected light curves and illustrates the varying cadences and coverage that different fields may have over the same one-year period. The PTF data are not publicly available, but ... TODO

@@

Applying the above prescription to the PTF data presents obvious challenges. For example, microlensing surveys have relatively uniform time sampling of their survey footprint over an observing season, justifying the first requirement. But it is harder to motivate such a cut on data with significant and irregular gaps. In order to develop the most successful procedure for PTF data, we therefore examine the relative performance of a set of variability indices and of the traditional prescription in selecting simulated microlensing events.

\begin{figure}[h]
\centering\includegraphics[width=.48\textwidth, trim=10 50 50 50, clip]{fig5.eps}
\caption{Light curves selected from PTF field 100101 with maximally outlying values for each variability index. For $\eta$ and $K$, this corresponds to the light curve with the \emph{minimum} value of the index over the entire field. For the other indices, these are the light curves with the \emph{maximum} value of the relevant index over the field. These light curves represent the type of variability that each index is most sensitive to. } \label{fig:indices_examples}
\end{figure}

\subsection{The Variability Indices} 
We choose five statistical measures of variability compiled by \cite{shin2009} that have been previously applied to the classification and discovery of periodic variables. These variability indices are $\sigma/\mu$, $Con$, $\eta$, $J$, and $K$\footnote{We do not implement the sixth index described by \cite{shin2009}, $AoVM$, because it mainly helps identify periodic sources.}. $\sigma/\mu$ is the ratio of the sample standard deviation to the sample mean, 
\begin{align}
	\frac{\sigma}{\mu} = \frac{\sqrt{\sum^N_i (x_i - \mu)^2 / (N-1)}}{\sum^N_i x_i/N},
\end{align}
where N is the total number of observations.

@@

$\eta$, the von Neumann ratio \citep[][also known as the Durbin-Watson statistic]{von_neumann1941}, is the mean square successive difference divided by the sample variance:
\begin{align}
	\eta = \frac{\delta^2}{\sigma^2} = \frac{\sum^{N-1}_i(x_{i+1} - x_i)^2/(N-1)}{\sigma^2}.
\end{align}
$\eta$ is small when there is strong positive serial correlation between successive data points.

@@

We add one more index, $\Delta \chi^2$, the difference in $\chi^2$ between fitting a Gaussian model and fitting a linear model to a light curve. In detail, we use a Levenberg-Marquardt optimizer to perform a least-squares fit with each of these models to the light curves and then compute $\Delta\chi^2=\chi_{linear} - \chi_{gaussian}$. Out tests with $\Delta \chi^2$ below compare the distribution of values over light curves on the same chip --- because of this, we neglect including an AIC or BIC correction to make this a proper likelihood test as the number of data points and number of model parameters are constant. For the Gaussian fit, the optimizer is initialized with a standard-deviation of 10 days, centered on the brightest data point. This is a standard statistical test used by microlensing surveys, and allows us to compare the relative performance of the (slightly modified) \cite{shin2009} indices and of this approach.

@@

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {fig7_b.eps}
	   \label{fig:subfig2}
	 }
\caption{{\it Top:} A randomly selected light curve from PTF field 4327. Note the sampling pattern. {\it Bottom:} Detection efficiency $\varepsilon$ for $\eta$, $\Delta\chi^2$, and $J$, as a function of the simulated microlensing event parameters $t_E$, $u_0$, and $m_0$ (the event timescale, impact parameter, and `quiescent' source magnitude, respectively). Black (dashed) lines show the distributions for all light curves (normalized, so scale of y-axis is arbitrary), red (solid) lines show the recovered distributions.}  \label{fig:detection_efficiency_4327}
\end{figure}

@@

\section{Searching for events in the full PTF dataset}\label{sec:search}
To identify candidate microlensing events in the $1.1\times10^9$ light curves with $>$10 $R$-band observations, we:
%We apply the 1\% FPR cut to the 2097 PTF fields ($\sim$15,224 deg$^2$) with $>$10 $R$-band observations using the variability index $\eta$. Our procedure is as follows:
\begin{enumerate}
	\item select a field from the 2097 fields with $>$10 good $R$-band observations and run Monte Carlo simulations to compute the 1\% FPR limiting values for $\eta$, $\Delta \chi^2$, and $J$ for that field;
	\item compute $\eta$ for all the light curves in the field, and select the light curves that pass the quality cut described in Section~\ref{sec:detection_eff} ($\sim$10\%) and the FPR cut determined from the Monte Carlo simulations ($\sim$0.1-0.5\%);
	\item for fields with SDSS coverage, we use Data Release 9 photometry \citep{dr9paper} to remove sources typed as Galaxy or QSO (using \texttt{objc\_type}\footnote{http://www.sdss3.org/dr9/algorithms/classify.php}) and flag candidate quasars \citep[using the cuts described in][]{richards02} ($\sim$10\%);
	\item fit a microlensing event model and subtract the model, then recompute $\eta$ for all the selected light curves;
	\item reject the light curves for which the new value of $\eta$ still passes the cut\footnote{These are probably periodic variables.} ($\sim$5-10\%);
	\item retain any remaining light curves for further inspection.
\end{enumerate}
% Turn this into a table, with number of objects left over after each cut?

This procedure identifies 2377 candidates from among the initial sample of $1.1\times10^9$ light curves. We search SIMBAD\footnote{This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France.} and remove any candidate among these with a known extragalactic counterpart within 10\asec. For those in the SDSS footprint, we examine the SDSS images \citep{sdss_images} to remove extended objects. Removing extragalactic objects in this manner from our candidate list reduces contamination by, e.g., supernovae.

@@

PTF1J0618 and PTF1J1728 have counterparts in 2MASS, and we can therefore use the \citet{kev07} $J$-band absolute magnitudes for stars of the relevant spectral types to estimate rough distances to these stars, assuming they have solar metallicity (see below for discussion of how we obtain spectral types for these stars). For PTF1J0618, this results in a distance that ranges from 1.1 kpc (if it is a M1 star) to 0.8 kpc (if it is a M3 star), corresponding to heights above the Galactic Plane 43 and 31 pc, respectively. For PTF1J1728, assuming it is a K0-K5 star results in distances ranging from 1.3 to 0.7 kpc, and heights of 0.9 to 0.5~kpc.

We lack a spectrum for PTF1J1615, so the distance estimate to this star is even more uncertain. Still, we compare the PTF $R$ magnitude to the stellar library compiled by \citet{Pickles1998} to estimate a distance, assuming (from its photometry) that it is a K5-M2 star\footnote{This ignores differences between Mould and Cousins $R$ filters.}. This results in distances ranging from 4.2 to 1.9 kpc and heights from 2.3 to 1.0~kpc. The estimated heights above the Galactic Plane suggest that PTF1J0618 is a thin disk star, while the other two, PTF1J1615 and PT1J1728, are plausible thick disk members \citep{Bochanski2010}.

@@

\subsection{Plausible Microlensing Candidates}
\subsubsection*{PTF1J061800.25$+$203142.5} % (a)}
%PTFS1206i
% - Star
% - 2MASS photometry believable
% - Detected in WISE -- appears blended with nearby object, photometry claims to have successfully deblended. 
PTF1J0618 is detected in 2MASS and in all four WISE filters. The corresponding colors are difficult to interpret, however: the object's $(J-H)$ color is consistent with that of a mid-M dwarf \citep{kev07}, but its $(H-K)$ is significantly redder than expected for such a star. Meanwhile, PTF1J0618's WISE colors suggest it is an extragalactic object \citep[cf.\ Figure 14 in][]{yan2013}. We observed PTF1J0618 for 1200~s with the Double Spectrograph \citep[DBSP;][]{oke_gunn} on the Hale 5-m telescope at Palomar Observatory, CA, on 2013 Feb 19; see Figure~\ref{fig:1206i_spectrum}. The spectrum was obtained with the D55 dichroic; from the atmospheric cutoff to 5500~\AA, the grating had 600 line~mm$^{-1}$ and was blazed at 4000~\AA, giving a resolution of 1.1~\AA. From 6300-8800~\AA, the grating was 158 line~mm$^{-1}$, was blazed at 7500~\AA, and gave a resolution of 2.5~\AA. The data were reduced using standard IRAF routines\footnote{IRAF is distributed by the National Optical Astronomy Observatories, which are operated by the Association of Universities for Research in Astronomy, Inc., under cooperative agreement with the National Science Foundation.}.

@@

Could PTF1J0618 be an outbursting symbiotic star? To test this hypothesis, we compare the DBSP spectrum to a set of M1-M2 dwarf and giant star spectra. We inspect the depth of the Ca II triplet lines and also calculate several of the spectral indices collected by \citet{kev06} to distinguish between dwarfs and giants (e.g., the fluxes in the Na D 5900 and CN 7900 bands). These all indicate that PTF1J0618 is a dwarf star, rendering it unlikely that the PTF event was a symbiotic-type outburst. We further note that those outbursts typically last $\sim$years \citep[e.g.,][]{tomov2013}. We conclude that this is a good candidate microlensing event.

@@

\subsubsection*{PTF1J161502.39$+$540053.8} % (c)}
%PTFS1216bt
PTF1J1615 is one of two candidates that fall in the SDSS footprint; its counterpart, SDSS J161502.43$+$540053.6, has a proper motion of 2.66 mas/year, consistent with its automated SDSS classification as a star. The star has $u = 22.9$ mag\footnote{These are SDSS PSF magnitudes. Typically, PSF fitting provides better estimates of isolated star magnitudes; see \citet{stoughton02}.}, beyond the $u = 22.0$ 95\% completeness limit for SDSS \citep{stoughton02}, and we therefore focus on its $griz$ photometry. With $(g-r) = 1.10\pm0.05$, $(r-i) = 1.00\pm0.03$, and $(i-z) = 0.56\pm0.04$, the star's colors are consistent with those of a late K/early M star \citep{kev07}. PTF1J1615 has no counterpart in 2MASS, and is only detected in the WISE $W1$ and $W2$ bands, making it hard to draw any conclusions based on these data. Still, its $(W1-W2)$ is consistent with what is seen for stars \citep[cf.\ Figure 14 in][]{yan2013}. Here again, the most likely source of contamination is flaring. However, the event detected by PTF had a duration $\sim$20 days, and so is highly unlikely to have been a flare. We conclude that this is a good candidate microlensing event.

@@

\subsubsection*{PTF1J193330.11$+$451501.1} % (z)}
%PTFS1219ad
PTF1J1933 is detected in both 2MASS and WISE. The 2MASS colors suggest that this is not a star \citep{kev07}, while the WISE colors are consistent with PTF1J1933 being an elliptical galaxy \citep[cf.\ Figure 14 in][]{yan2013}. This interpretation is strengthened by a visual inspection of the PTF and WISE images, which show that the object is elongated in the Northeast-Southwest direction. We tentatively conclude that this not a good microlensing candidate.

@@

We found that the von Neumann ratio, $\eta$ \citep{von_neumann1941}, performs better than previously used statistics in recovering injected microlensing events in non-uniformly sampled data. We used $\eta$ to develop a selection procedure for extracting microlensing event candidates from the PTF light-curve archive. Among the $1.1 \times 10^9$ light curves with $>$10 $R$-band observations, we first identified $\sim$2000 interesting candidate transients, from which we selected 11 candidate microlensing events.

>>>>

%\slugcomment{DRAFT \today}
\shorttitle{Microlensing \& PTF}
\shortauthors{Price-Whelan et al.}

@@

\title{Statistical Searches for Microlensing Events in Large, Non-Uniformly Sampled Time-Domain Surveys: A Test Using Palomar Transient Factory Data}
\author{Adrian~M.~Price-Whelan\altaffilmark{1} Marcel~A.~Ag\"ueros\altaffilmark{1}, Amanda P. Fournier\altaffilmark{2,3}, Rachel Street\altaffilmark{3}, Eran O. Ofek\altaffilmark{4}, Kevin~R.~Covey\altaffilmark{5}, David Levitan\altaffilmark{6}, Russ R.~Laher\altaffilmark{7}, Branimir Sesar\altaffilmark{6}, Jason Surace\altaffilmark{7}}

\altaffiltext{1}{Department of Astronomy, Columbia University, 550 W 120th St., New York, NY 10027, USA; adrn@astro.columbia.edu}
\altaffiltext{2}{Department of Physics, Broida Hall, University of California, Santa Barbara, CA 93106, USA}
\altaffiltext{3}{Las Cumbres Observatory Global Telescope Network, Inc., 6740 Cortona Dr.\ Suite 102, Santa Barbara, CA 93117, USA}
\altaffiltext{4}{Benoziyo Center for Astrophysics, Weizmann Institute of Science, 76100 Rehovot, Israel}
\altaffiltext{5}{Lowell Observatory, 1400 West Mars Hill Rd, Flagstaff, AZ 86001, USA}
\altaffiltext{6}{Division of Physics, Mathematics, and Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
\altaffiltext{7}{Spitzer Science Center, California Institute of Technology, Mail Stop 314-6, Pasadena, CA 91125, USA}

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged $\sim$40 times in \textit{R}-band, $\sim$$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal time sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio, also known as the Durbin-Watson statistic, performs best for identifying simulated microlensing events in our data. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations --- a total of $1.1\times10^9$ light curves --- to uncover three candidate sparse-field microlensing events. These candidates cannot be confirmed, as we lack simultaneous, multi-color photometry of the events. However, their number is consistent with predictions for the microlensing event rate in the PTF footprint over the survey's three years of operations, as estimated from near-field microlensing models. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large data sets, both of which will be useful to future wide-field, time-domain surveys, such as that planned with the Large Synoptic Survey Telescope. 
\end{abstract}

@@

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005}, and to identify extrasolar planets \citep[][and references therein]{gaudi2011}. Such studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year.\footnote{e.g., {\tt http://ogle.astrouw.edu.pl/ogle4/ews/ews.html}} In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects, such as low-mass stars, sub-stellar objects, and isolated neutron stars.

Microlensing surveys typically focus on high-density stellar regions, such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.,][]{original_ogle, original_macho, eros_original, crotts1996}, but microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the microlensing event rate in low-density stellar regions will be smaller, compared to events found by modern microlensing surveys, these are expected to involve closer sources and lenses, and less crowded backgrounds \citep{mesolensing}. Thus, such lenses offer more opportunities to constrain the properties of the events.

@@

% TODO: Han citation above, also mentioned in one of Rosanne's papers

In this paper, we use the Palomar Transient Factory (PTF) survey, described in Section~\ref{sec:ptf}, as a test case for studying the frequency and detectability of microlensing events in all-sky, irregularly sampled, time-domain surveys. Although to date no microlensing event has been reported using PTF data, a simple extrapolation from previous all-sky rate estimates suggests $\sim$1--10 detectable microlensing events within the PTF footprint over the survey's three years of operation \citep{gaudi2008, han2008}. We examine the selection methods used by microlensing surveys to identify event signatures in light curves as well as an event identification procedure based on a set of variability statistics. We then present the detection efficiency for these methods as computed for a representative set of PTF fields (Section~\ref{sec:event_recovery}). In Section~\ref{sec:search}, we apply a set of criteria that use the results of these tests to all PTF light curves with $>$10 high-quality observations to identify candidate microlensing events. We discuss eight interesting transients and three candidate events identified in this manner before concluding in Section~\ref{sec:conc}. % A detailed investigation of the all-sky microlensing event rate,  generated from a model that includes realistic mass and velocity distributions, will be presented in Fournier et al. (in prep).

@@

As of 2012 Dec, the PTF footprint includes $\sim$15,224 (2766) deg$^2$ imaged $>$10 ($>$100) times in $R$-band and $\sim$5430 (290) deg$^2$ imaged that often in $g$-band (see Figure~\ref{fig:survey_footprint}). The PTF survey footprint is not uniformly sampled either spatially or temporally. Each field has a unique sampling pattern determined by which of the PTF sub-surveys it belongs to, what time of the year it is visible, and how high a priority it is given by the scheduler. The imaging cadences range from 1-5 days, but there is a higher-cadence campaign of a field in Orion to find transiting planets around young stars \citep{nick2009}. The result is that PTF light curves often contain gaps and regions of high-cadence observations and/or of low-cadence observations, so that the archive is a massive dataset of irregularly-sampled, time-domain photometry. Figure~\ref{fig:sampling} shows six randomly selected light curves and illustrates the varying cadences and coverage that different fields may have over the same one-year period. While the PTF database is not yet public, we have made $10^5$ light curves available at {\tt }. These light curves meet the quality criteria described in Section~\ref{sec:detection_eff} and can be used to test the statistical methods described here (or others!).

@@

Applying the standard microlensing-search prescription to the PTF data presents obvious challenges. For example, microlensing surveys have relatively uniform time sampling of their survey footprint over an observing season, justifying the first requirement. But it is harder to motivate such a cut on data with significant and irregular gaps. In order to develop the most successful procedure for PTF data, we therefore examine the relative performance of a set of variability indices and of the traditional prescription in selecting simulated microlensing events.

\begin{figure}[h]
\centering\includegraphics[width=.48\textwidth, trim=10 50 50 50, clip]{fig5.eps}
\caption{Light curves selected from PTF field 3756 with maximally outlying values for each variability index. For $\eta$ and $K$, this corresponds to the light curve with the \emph{minimum} value of the index over the entire field. For the other indices, these are the light curves with the \emph{maximum} value of the relevant index over the field. These light curves illustrate the type of variability that each index is most sensitive to. } \label{fig:indices_examples}
\end{figure}

\subsection{The Variability Indices} 
We choose five statistical measures of variability compiled by \cite{shin2009} that have been previously applied to the classification and discovery of periodic variables. These variability indices are $\sigma/\mu$, $Con$, $\eta$, $J$, and $K$.\footnote{We do not implement the sixth index described by \cite{shin2009}, $AoVM$, because it mainly helps identify periodic sources.} $\sigma/\mu$ is the ratio of the sample standard deviation ($\sigma$) to the sample mean, 
\begin{align}
	\frac{\sigma}{\mu} = \frac{\sqrt{\sum^N_i (x_i - \mu)^2 / (N-1)}}{\sum^N_i x_i/N},
\end{align}
where N is the total number of observations.

@@

$\eta$, the von Neumann ratio \citep[also known as the Durbin-Watson statistic;][]{von_neumann1941, durbin50}, is the mean square successive difference divided by the sample variance:
\begin{align}
	\eta = \frac{\delta^2}{\sigma^2} = \frac{\sum^{N-1}_i(x_{i+1} - x_i)^2/(N-1)}{\sigma^2}.
\end{align}
$\eta$ is small when there is strong positive serial correlation between successive data points.

@@

We add one more index, $\Delta \chi^2$, the difference in $\chi^2$ between fitting a Gaussian model and fitting a linear model to a light curve. Specifically, we use a Levenberg-Marquardt optimizer to perform a least-squares fit with each of these models to the light curves and then compute $\Delta\chi^2=\chi_{linear} - \chi_{gaussian}$. Our tests with $\Delta \chi^2$ below compare the distribution of values over light curves on the same chip. As a result, the number of data points and number of model parameters are constant, and we therefore do not include an Akaike or Bayesian information criterion term.

For the Gaussian fit, the optimizer is initialized with a $\sigma$ of 10 days, centered on the brightest data point. This is a standard statistical test used by microlensing surveys, and allows us to compare the relative performance of the (slightly modified) \cite{shin2009} indices and of this approach.

@@

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {fig7_b.eps}
	   \label{fig:subfig2}
	 }
\caption{{\it Top:} A randomly selected light curve from PTF field 4327. Note the sampling pattern. {\it Bottom:} Detection efficiency $\varepsilon$ for $\eta$, $\Delta\chi^2$, and $J$, as a function of the simulated microlensing event parameters $t_E$, $u_0$, and $m_0$ (the event timescale, impact parameter, and unmagnified source magnitude, respectively). Black (dashed) lines show the distributions for all light curves (normalized, so y-axis scale is arbitrary), red (solid) lines show the recovered distributions.}  \label{fig:detection_efficiency_4327}
\end{figure}

@@

\section{Searching for events in the full PTF dataset}\label{sec:search}
To identify candidate microlensing events in the $1.1\times10^9$ light curves with $>$10 $R$-band observations, we:
%We apply the 1\% FPR cut to the 2097 PTF fields ($\sim$15,224 deg$^2$) with $>$10 $R$-band observations using the variability index $\eta$. Our procedure is as follows:
\begin{enumerate}
	\item select a field from the 2097 fields with $>$10 good $R$-band observations and run Monte Carlo simulations to compute the 1\% FPR limiting values for $\eta$, $\Delta \chi^2$, and $J$ for that field;
	\item compute $\eta$ for all the light curves in the field. While $\sim$10\% of the light curves pass the quality cut described in Section~\ref{sec:detection_eff}, only $\sim$0.1-0.5\% of these pass the FPR cut determined from the Monte Carlo simulations;
	\item for fields with SDSS coverage, we use Data Release 9 photometry \citep{dr9paper} to remove sources typed as Galaxy or QSO (using \texttt{objc\_type}\footnote{{\tt http://www.sdss3.org/dr9/algorithms/classify.php}}) and flag candidate quasars \citep[using the cuts described in][]{richards02} ($\sim$10\%);
	\item fit a microlensing event model and subtract the model, then recompute $\eta$ for all the selected light curves;
	\item reject the light curves for which the new value of $\eta$ still passes the cut\footnote{These are probably periodic variables.} ($\sim$5-10\%);
	\item retain any remaining light curves for further inspection.
\end{enumerate}
% Turn this into a table, with number of objects left over after each cut?

This procedure identifies 2377 candidates from among the initial sample of $1.1\times10^9$ light curves. We search SIMBAD\footnote{This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France.} and remove any candidate among these with a known extragalactic counterpart within 10\asec. For those in the SDSS footprint, we examine the SDSS images \citep{sdss_images} to identify extended objects. Removing extragalactic objects in this manner from our candidate list reduces contamination by, e.g., supernovae.

@@

PTF1J0618 and PTF1J1728 have counterparts in 2MASS, and we can therefore use the \citet{kev07} $J$-band absolute magnitudes for stars of the relevant spectral types to estimate rough distances to these stars, assuming they have solar metallicity (see below for discussion of how we obtain spectral types for these stars). For PTF1J0618, this results in a distance that ranges from 1.1 kpc (if it is a M1 star) to 0.8 kpc (if it is a M3 star), corresponding to heights above the Galactic Plane of 43 and 31 pc, respectively. For PTF1J1728, assuming it is a K0-K5 star results in distances ranging from 1.3 to 0.7 kpc, and heights of 0.9 to 0.5~kpc.

We lack a spectrum for PTF1J1615, so the distance estimate to this star is even more uncertain. Still, we compare the PTF $R$ magnitude to the stellar library compiled by \citet{Pickles1998} to estimate a distance, assuming (from its photometry) that it is a K5-M2 star.\footnote{This ignores differences between Mould and Cousins $R$ filters.} This results in distances ranging from 4.2 to 1.9 kpc and heights from 2.3 to 1.0~kpc. The estimated heights above the Galactic Plane suggest that PTF1J0618 is a thin disk star, while the other two, PTF1J1615 and PT1J1728, are plausible thick disk members \citep{Bochanski2010}.

@@

\subsection{Plausible Microlensing Candidates}
\subsubsection*{PTF1J061800.25$+$203142.5} % (a)}
%PTFS1206i
% - Star
% - 2MASS photometry believable
% - Detected in WISE -- appears blended with nearby object, photometry claims to have successfully deblended. 
PTF1J0618 is detected in 2MASS and in all four WISE filters. The corresponding colors are difficult to interpret, however: the object's $(J-H)$ color is consistent with that of a mid-M dwarf \citep{kev07}, but its $(H-K)$ is significantly redder than expected for such a star. Meanwhile, PTF1J0618's WISE colors suggest it is an extragalactic object \citep[cf.\ Figure 14 in][]{yan2013}. We observed PTF1J0618 for 1200~s with the Double Spectrograph \citep[DBSP;][]{oke_gunn} on the Hale 5-m telescope at Palomar Observatory, CA, on 2013 Feb 19; see Figure~\ref{fig:1206i_spectrum}. The spectrum was obtained with the D55 dichroic; from the atmospheric cutoff to 5500~\AA, the grating had 600 line~mm$^{-1}$ and was blazed at 4000~\AA, giving a resolution of 1.1~\AA. From 6300-8800~\AA, the grating was 158 line~mm$^{-1}$, was blazed at 7500~\AA, and gave a resolution of 2.5~\AA. The data were reduced using standard IRAF routines.\footnote{IRAF is distributed by the National Optical Astronomy Observatories, which are operated by the Association of Universities for Research in Astronomy, Inc., under cooperative agreement with the National Science Foundation.}

@@

Could PTF1J0618 be an outbursting symbiotic star? To test this hypothesis, we compare the DBSP spectrum to a set of M1-M2 dwarf and giant star spectra. We inspect the depth of the Ca II triplet lines and also calculate several of the spectral indices collected by \citet{kev06} to distinguish between dwarfs and giants (e.g., the fluxes in the Na D 5900 and CN 7900 bands). These all indicate that PTF1J0618 is a dwarf star, rendering it unlikely that the PTF event was a symbiotic-type outburst. We further note that those outbursts typically last $\sim$years \citep[e.g.,][]{tomov2013}. We conclude that this is a plausible candidate microlensing event.

@@

\subsubsection*{PTF1J161502.39$+$540053.8} % (c)}
%PTFS1216bt
PTF1J1615 is one of two candidates that fall in the SDSS footprint; its counterpart, SDSS J161502.43$+$540053.6, has a proper motion of 2.66 mas/year, consistent with its automated SDSS classification as a star. The star has $u = 22.9$ mag,\footnote{These are SDSS PSF magnitudes. Typically, PSF fitting provides better estimates of isolated star magnitudes; see \citet{stoughton02}.} beyond the $u = 22.0$ 95\% completeness limit for SDSS \citep{stoughton02}, and we therefore focus on its $griz$ photometry. With $(g-r) = 1.10\pm0.05$, $(r-i) = 1.00\pm0.03$, and $(i-z) = 0.56\pm0.04$, the star's colors are consistent with those of a late K/early M star \citep{kev07}. PTF1J1615 has no counterpart in 2MASS, and is only detected in the WISE $W1$ and $W2$ bands, making it hard to draw any conclusions based on these data. Still, its $(W1-W2)$ is consistent with what is seen for stars \citep[cf.\ Figure 14 in][]{yan2013}. Here again, the most likely source of contamination is flaring. However, the event detected by PTF had a duration $\sim$20 days, and so is highly unlikely to have been a flare. We conclude that this is a plausible candidate microlensing event.

@@

\subsubsection*{PTF1J193330.11$+$451501.1} % (z)}
%PTFS1219ad
PTF1J1933 is detected in both 2MASS and WISE. The 2MASS colors suggest that this is not a star \citep{kev07}, while the WISE colors are consistent with PTF1J1933 being an elliptical galaxy \citep[cf.\ Figure 14 in][]{yan2013}. This interpretation is strengthened by a visual inspection of the PTF and WISE images, which show that the object is elongated in the Northeast-Southwest direction. We tentatively conclude that this not a plausible microlensing candidate.

@@

We found that the von Neumann ratio, $\eta$ \citep[or Durbin-Watson statistic;][]{von_neumann1941, durbin50}, performs better than previously used statistics in recovering injected microlensing events in non-uniformly sampled data. We used $\eta$ to develop a selection procedure for extracting microlensing event candidates from the PTF light-curve archive. Among the $1.1 \times 10^9$ light curves with $>$10 $R$-band observations, we first identified $\sim$2000 interesting candidate transients, from which we selected 11 candidate microlensing events.
